Efficient pedigree recording for fast population genetics simulation

### 2018-01-12 forwards_paper.tex
succinct ts as in #54

\begin{abstract}
    In this paper we describe how to
    efficiently record the entire genetic history of a population
    during a forwards-time, individual-based population genetics simulation.
    This dramatically reduces the computational burden of tracking individual genomes
    by simulating only those loci that may affect reproduction (those having non-neutral variants),
    and recording history in the \emph{tree sequence} format introduced in the software package \msprime,
    on which neutral mutations can be quickly placed afterwards.
    We implement the method in two popular forwards-time simulation frameworks,
    which speeds up large, whole-genome simulations by one or two orders of magnitude.
    In addition to speed, we show that the method has several advantages:
    (1) All marginal genealogies of the simulated individuals are recorded, rather than just genotypes.
    (2) A population of $N$ individuals with $M$ polymorphic sites
    can be stored on $O(N \log N + M)$ space, making it feasible to store a simulation's entire final generation
    rather than only a subset.
    (3) A simulation can easily be initialized with prior history produced by a more efficient coalescent simulation of deep history.
    To make these possible,
    we introduce the \msprime{} Tables API, which allows for the efficient interchange of genealogical data between modular program components.
\end{abstract}

@@

However, it is thought that most genetic variation is selectively neutral (or nearly so).
By definition, the neutral alleles carried by individuals in a population
do not affect the population process.
For this reason, if one records the entire genealogical history of a population over the course of a simulation,
simply laying down neutral mutations on top of that history afterwards
is equivalent to having generated them during the simulation.
Precisely, we would need to know the genealogical tree relating all sampled individuals
at each position along the genome.
This is equivalent to knowing the \emph{population pedigree} --
the complete history of parent-offspring relationships of an entire population
going back to a remote time -- and the genetic outcomes of each ancestral
meiosis.
The embellished pedigree contains all the information necessary
to construct the genealogical tree that relates each individual to every other
at each position on the genome,
i.e., the \emph{tree sequence} implemented in the coalescent simulator \msprime{}~\citep{kelleher2016efficient}.
Combined with ancestral genotypes and the origins of new mutations,
it also completely specifies the genomic sequence of any individual in the population at any time.
This is much much more than typically recorded in a forward simulation, however,
so further savings can be had by discarding all information irrelevant to the genetic history
of the extant population.
\jda{The space complexity result seemed like a result in the intro; and maybe didn't include some detail
	about assumed number of generations}

In this paper, we describe storage methods for genealogies
(and hence, genome sequence) and an algorithm for simplifying these,
both implemented as improvements to the algorithmic tools and data structures in \msprime{}.
We also describe a general method of using these tools to efficiently record,
and later process, the population pedgiree in forwards-time simulation.
While providing substantial savings in time and space, our methods provide much
more information than simply simulating the genomes -- the tree sequence format
encodes all marginal genealogies of individuals living at the end of the
simulation.
Although we were motivated by a need for more efficient genomic simulations,
these tools may prove more widely useful.

>>>>

\begin{abstract}
    In this paper we describe how to
    efficiently record the entire genetic history of a population
    during a forwards-time, individual-based population genetics simulation.
    This dramatically reduces the computational burden of tracking individual genomes
    by simulating only those loci that may affect reproduction (those having non-neutral variants),
    and recording history as a \emph{succinct tree sequence} as introduced in the software package \msprime,
    on which neutral mutations can be quickly placed afterwards.
    We implement the method in two popular forwards-time simulation frameworks,
    which speeds up large, whole-genome simulations by one or two orders of magnitude.
    In addition to speed, we show that the method has several advantages:
    (1) All marginal genealogies of the simulated individuals are recorded, rather than just genotypes.
    (2) A population of $N$ individuals with $M$ polymorphic sites
    can be stored on $O(N \log N + M)$ space, making it feasible to store a simulation's entire final generation
    rather than only a subset.
    (3) A simulation can easily be initialized with prior history produced by a more efficient coalescent simulation of deep history.
    To make these possible,
    we introduce the \msprime{} Tables API, which allows for the efficient interchange of genealogical data between modular program components.
\end{abstract}

@@

However, it is thought that most genetic variation is selectively neutral (or nearly so).
By definition, the neutral alleles carried by individuals in a population
do not affect the population process.
For this reason, if one records the entire genealogical history of a population over the course of a simulation,
simply laying down neutral mutations on top of that history afterwards
is equivalent to having generated them during the simulation.
Precisely, we would need to know the genealogical tree relating all sampled individuals
at each position along the genome.
This is equivalent to knowing the \emph{population pedigree} --
the complete history of parent-offspring relationships of an entire population
going back to a remote time -- and the genetic outcomes of each ancestral meiosis.
The embellished pedigree contains all the information necessary
to construct the genealogical tree that relates each individual to every other
at each position on the genome.
Combined with ancestral genotypes and the origins of new mutations,
it also completely specifies the genomic sequence of any individual in the population at any time.
This is much much more than typically recorded in a forward simulation, however,
so further savings can be had by discarding all information irrelevant to the genetic history
of the extant population.

In this paper, we describe a method for storing a \emph{succinct tree sequence} 
(or, for brevity, ``tree sequence''),
and, hence, genome sequence, as well as an algorithm for simplifying these.
Both are implemented as improvements to the algorithmic tools and data structures in \msprime{}.
We also describe a general method of using these tools to efficiently record,
and later process, the population pedgiree in forwards-time simulation.
While providing substantial savings in time and space, our methods provide much
more information than simply simulating the genomes -- the tree sequence
encodes all marginal genealogies of individuals living at the end of the simulation.
Although we were motivated by a need for more efficient genomic simulations,
these tools may prove more widely useful.
